---
title: "R Notebook"
output: html_notebook
---

# Libraries

```{r}
library(data.table)
library(tidyverse)
library(readxl)
library(ggVennDiagram)
library(maftools)
```


# Download MAF
First we download the PCAWG final SNV,MNV,INDEL calls. This file contains the ICGC portion of the PCAWG consensus callsets for SNV/Indel.

`https://dcc.icgc.org/api/v1/download?fn=/PCAWG/consensus_snv_indel/final_consensus_passonly.snv_mnv_indel.icgc.public.maf.gz`

```{r eval=FALSE, include=FALSE}
pcawg_maf_url = "https://dcc.icgc.org/api/v1/download?fn=/PCAWG/consensus_snv_indel/final_consensus_passonly.snv_mnv_indel.icgc.public.maf.gz"
pcawg_maf_path = pcawg_maf_url %>% strsplit(split = "/") %>% unlist %>% tail(n=1)

download.file(pcawg_maf_url, destfile = pcawg_maf_path) # commented to make sure it never runs
```


If you don't want MNVs and just want SNVs and indels, download:
`https://dcc.icgc.org/api/v1/download?fn=/PCAWG/consensus_snv_indel/final_consensus_snv_indel_passonly_icgc.public.tgz`

See `https://dcc.icgc.org/releases/PCAWG/consensus_snv_indel` for details.


# Prepare MAF data

```{r eval=FALSE, include=FALSE}
pcawg_df <- data.table::fread("final_consensus_passonly.snv_mnv_indel.icgc.public.maf")
pcawg_df
```

Next we select only the most important MAF columns to save on space

```{r eval=FALSE, include=FALSE}
pcawg_minimal_columns_df <- pcawg_df %>% dplyr::select("Hugo_Symbol", "Chromosome", Start_Position = "Start_position", End_Position = "End_position", "Reference_Allele", "Tumor_Seq_Allele2", "Variant_Classification", "Variant_Type", "Tumor_Sample_Barcode", "t_ref_count", "t_alt_count", "Donor_ID","Project_Code")
```


We'll then split data by project_code, so we can create a separate MAF object for each cohort

```{r eval=FALSE, include=FALSE}

# Look at all the different Project Codes
pcawg_minimal_columns_df %>%
  dplyr::count(Project_Code, sort = TRUE)

# Split df into a list of dfs, one for each Project_Code
pcawg_split_ls = split(pcawg_minimal_columns_df, f = pcawg_minimal_columns_df$Project_Code)
```


# Prepare Metadata

Before we create our MAF objects, we'll need some metadata:

1. Download histology metadata from
`https://dcc.icgc.org/api/v1/download?fn=/PCAWG/clinical_and_histology/pcawg_specimen_histology_August2016_v9.xlsx`

2. Download clinical metadata from 
`https://dcc.icgc.org/api/v1/download?fn=/PCAWG/clinical_and_histology/pcawg_donor_clinical_August2016_v9.xlsx`

3. We'll also need the sample sheet to bring everything together:
 `https://dcc.icgc.org/api/v1/download?fn=/PCAWG/data_releases/latest/pcawg_sample_sheet.tsv`
This `aliquot_id` column in the sample sheet corresponds to the 'Tumor_Sample_Barcodes' in the MAF

See
`https://dcc.icgc.org/releases/PCAWG/clinical_and_histology` and `https://dcc.icgc.org/releases/PCAWG/data_releases/latest` for more info

## Download
```{r eval=FALSE, include=FALSE}
hist_url = "https://dcc.icgc.org/api/v1/download?fn=/PCAWG/clinical_and_histology/pcawg_specimen_histology_August2016_v9.xlsx"
hist_path = hist_url %>% strsplit(split = "/") %>% unlist %>% tail(n=1)

clin_url = "https://dcc.icgc.org/api/v1/download?fn=/PCAWG/clinical_and_histology/pcawg_donor_clinical_August2016_v9.xlsx"
clin_path = clin_url %>% strsplit(split = "/") %>% unlist %>% tail(n=1)

sample_sheet_url = "https://dcc.icgc.org/api/v1/download?fn=/PCAWG/data_releases/latest/pcawg_sample_sheet.tsv"
sample_sheet_path = sample_sheet_url %>% strsplit(split = "/") %>% unlist %>% tail(n=1)


download.file(hist_url, destfile = hist_path)
download.file(clin_url, destfile = clin_path)
download.file(sample_sheet_url, destfile = sample_sheet_path)
```


## Reformat

```{r eval=FALSE, include=FALSE}

# Read data
hist_df <- read_excel(path = hist_path)
clin_df <- read_excel(path = clin_path)
sample_df <- fread(sample_sheet_path)

sample_level_metadata <- full_join(hist_df, clin_df, by = "icgc_donor_id")
sample_level_metadata <- full_join(sample_df, sample_level_metadata, by = "icgc_sample_id") %>% 
  filter(!is.na(aliquot_id))

if(anyDuplicated(sample_level_metadata$aliquot_id) > 0)
  message("PROBLEM WITH METADATA, DUPLICATED ALIQUOT ID")

sample_level_metadata <- sample_level_metadata %>% 
  dplyr::rename(Tumor_Sample_Barcode = aliquot_id)

```


# Create MAFs

Ensure metadata contains all Tumor_Sample_Barcodes in maf dataframe
```{r}

sample_level_metadata <- sample_level_metadata %>% filter(Tumor_Sample_Barcode %in% pcawg_minimal_columns_df$Tumor_Sample_Barcode)

 ggVennDiagram(list(
   maf=unique(pcawg_minimal_columns_df$Tumor_Sample_Barcode), 
   metadata=sample_level_metadata$Tumor_Sample_Barcode))

```

```{r}
outfolder = "rds"

dir.create(outfolder)

pcawg_split_ls %>% 
  seq_along %>% 
  lapply(function(n) {
    name = names(pcawg_split_ls)[n] %>% gsub(" ", "_", x = .)
    maf = read.maf(maf = pcawg_split_ls[[n]], clinicalData = sample_level_metadata)
    saveRDS(object = maf, file = paste0(outfolder, "/", name, ".rds"))
  } )

```

# Miscellaneous 
For more info on what disease-types were binned into each of the PCAWG projects, download tumor map.

```{r}
subtype_map_url = "https://dcc.icgc.org/api/v1/download?fn=/PCAWG/clinical_and_histology/tumour_subtype_consolidation_map.xlsx"
subtype_map_path = subtype_map_url %>% strsplit(split = "/") %>% unlist() %>% tail(1)

download.file(subtype_map_url, subtype_map_path)
```


